A look into Kamala Harris' polls in the upcoming 2024 US Presidential Election 

### 2024-11-04 paper/paper.qmd
Update paper.qmd
Added Data section, which includes table and graphs of relevant variables of discussion.

The data used in this model comes from fivethirtyeight.com, a polling site known for their in-depth analysis and data on US elections @fivethirtyeight. Here, we have used the 2024 general election data, originally accessed on 10/21/2024. The data contains information from several polls taken across the United States, including the operator, sponsors, methodology, a numeric grade to quanitfy the quality of the poll, and when and where the poll was taken. From this dataset, polls predicting a Harris win were considered, to construct the predictive model.

>>>>

The dataset provided by fivethirtyeight contains over 15000 observations across 50 variables. Each observation is a poll conducted on the 2024 US Presidential Election. Fivethirtyeight's dataset was chosen due to it's comprehensive review of polls and pollsters, attributing grades in error and bias, as well as in transparency. Their thorough investigations ensure their dataset not only contains insightful polls, but also includes as much relevant information about the polls conducted as possible. This gives us plenty of predictors when building a model that will caluclate the percentage win of either Trump or Harris.

The dataset was read into R [@citeR], and cleaned using the R package tidyverse [@tidyverse]. Due to Kamal's late entry into the race, some polls do not have any data on her specifically. This is why we set a cut-off date at 2024/07/21, the day Kamala replaced Joe Biden as the Democratic nominee.

Fivethirtyeight includes a metric called `numeric_grade`, which combines the transparency_score and transparency score. We only included polls with a `numeric_grade` at or above 2.7. This gave us the best balance between quantity and quality. Finally, we create 2 datasets, one with polls predicting a Harris win, the other a Trump win.

#### Create end_date - pct base plots for Trump and Harris ####
base_plot_kamala <- ggplot(just_harris_high_quality, aes(x = end_date, y = pct)) +
  theme_classic() +
  labs(y = "Harris percent", x = "Date")

base_plot_trump <- ggplot(just_trump_high_quality, aes(x = end_date, y = pct)) +
  theme_classic() +
  labs(y = "Trump percent", x = "Date")
```

## Predictors

### **`transparency_score`**

Fivethirtyeight awards a score out of 10 to each pollster based on how informative their methodology is. Understanding the methodology of a pollster assits in understanding any errors or biases present in their findings.

```{r}
#| label: fig-transscore_harris
#| fig-cap: As election day draws close, pollsters are wrapping up their polls and posting their predictions on who will win. The polls in this graph predict Harris' win percentage, with the average in the blue line. The dots in the graph below represent individual polls conducted in the months leading up to the election. They are coloured based on their transparency_score, which is fivethirtyeight's metric for transpareny in the pollster's methodology. 
#| echo: false
#| warning: false
#| message: false

# Color by transparency_score
base_plot_kamala +
  geom_point(aes(color = factor(transparency_score))) +
  geom_smooth(color = "blue") +
  theme(legend.position = "bottom")
```

```{r}
#| label: tbl-transscore_harris
#| tbl-cap: Summary of transparency_scores amongst polls predicting a Kamala Harris win. The closer the values are to 10, the more trustworthy the predictions. 
#| echo: false
#| warning: false
#| message: false

# Generating summary statistics for Harris
transparency_score_harris_summary <- just_harris_high_quality %>%
  summarize(
    Min = min(transparency_score, na.rm = TRUE),
    Mean = mean(transparency_score, na.rm = TRUE),
    Median = median(transparency_score, na.rm = TRUE),
    Max = max(transparency_score, na.rm = TRUE)
  )

# Displaying summary
transparency_score_harris_summary %>% 
  kable() %>%
  kable_styling()
```

```{r}
#| label: fig-transscore_trump
#| fig-cap: As election day draws close, pollsters are wrapping up their polls and posting their predictions on who will win. The polls in this graph predict Trump's win percentage, with the average in the blue line. The dots in the graph below represent individual polls conducted in the months leading up to the election. They are coloured based on their transparency_score, which is fivethirtyeight's metric for transpareny in the pollster's methodology.
#| echo: false
#| warning: false
#| message: false

# Color by transparency_score
base_plot_trump +
  geom_point(aes(color = factor(transparency_score))) +
  geom_smooth(color = "red") +
  theme(legend.position = "bottom")
```

```{r}
#| label: tbl-transscore_trump
#| tbl-cap: Summary of transparency_scores amongst polls predicting a Donald Trump win. The closer the values are to 10, the more trustworthy the predictions. 
#| echo: false
#| warning: false
#| message: false

# Generating summary statistics for Trump
numeric_grade_trump_summary <- just_trump_high_quality %>%
  summarize(
    Min = min(transparency_score, na.rm = TRUE),
    Mean = mean(transparency_score, na.rm = TRUE),
    Median = median(transparency_score, na.rm = TRUE),
    Max = max(transparency_score, na.rm = TRUE)
  )

# Display summary
transparency_score_trump_summary %>% 
  kable() %>%
  kable_styling()

### **`numeric_grade`**

A cumulative score given to pollster that combines fivethirtyeight's `pollscore` and `transparency_score`. Like `transparency_score`, we believe this metric can give us insight into potential biases in each pollster and poll, which is why it is included in our model.

```{r}
#| label: fig-numgrade_harris
#| fig-cap: As election day draws close, pollsters are wrapping up their polls and posting their predictions on who will win. The polls in this graph predict Harris' win percentage, with the average in the blue line. The dots in the graph below represent individual polls conducted in the months leading up to the election. They are coloured based on their numeric_grade, which is fivethirtyeight's combined metric on transparency and bias/errors. 
#| echo: false
#| warning: false
#| message: false

# Color by transparency_score
base_plot_kamala +
  geom_point(aes(color = factor(numeric_grade))) +
  geom_smooth(color = "blue") +
  theme(legend.position = "bottom")
```
```{r}
#| label: tbl-numgrade_harris
#| tbl-cap: Summary of numeric_grades amongst polls predicting a Kamala Harris win.
#| echo: false
#| warning: false
#| message: false

# Generating summary statistics for Harris
numeric_grade_harris_summary <- just_harris_high_quality %>%
  summarize(
    Min = min(numeric_grade, na.rm = TRUE),
    Mean = mean(numeric_grade, na.rm = TRUE),
    Median = median(numeric_grade, na.rm = TRUE),
    Max = max(numeric_grade, na.rm = TRUE)
  )

# Display summary
numeric_grade_harris_summary %>% 
  kable() %>%
  kable_styling()

```{r}
#| label: fig-numgrade_trump
#| fig-cap: As election day draws close, pollsters are wrapping up their polls and posting their predictions on who will win. The polls in this graph predict Trumps' win percentage, with the average in the blue line. The dots in the graph below represent individual polls conducted in the months leading up to the election. They are coloured based on their numeric_grade, which is fivethirtyeight's combined metric on transparency and bias/errors. 
#| echo: false
#| warning: false
#| message: false

# Color by transparency_score
base_plot_kamala +
  geom_point(aes(color = factor(numeric_grade))) +
  geom_smooth(color = "red") +
  theme(legend.position = "bottom")
```
```{r}
#| label: tbl-transscore_trump
#| tbl-cap: Summary of transparency_scores amongst polls predicting a Donald Trump win. The closer the values are to 10, the more trustworthy the predictions. 
#| echo: false
#| warning: false
#| message: false

# Generating summary statistics for Trump
numeric_grade_trump_summary <- just_trump_high_quality %>%
  summarize(
    Min = min(transparency_score, na.rm = TRUE),
    Mean = mean(transparency_score, na.rm = TRUE),
    Median = median(transparency_score, na.rm = TRUE),
    Max = max(transparency_score, na.rm = TRUE)
  )

# Display summary
numeric_grade_trump_summary %>% 
  kable() %>%
  kable_styling()

### `pollster`

```{r}
#| label: fig-pollster_harris
#| fig-cap: As election day draws close, pollsters are wrapping up their polls and posting their predictions on who will win. The polls in this graph predict Harris' win percentage, with the average in the blue line. The dots in the graph below represent individual polls conducted in the months leading up to the election. They are coloured based on the pollster who authored the poll. 
#| echo: false
#| warning: false
#| message: false

# Color by transparency_score
base_plot_kamala +
  geom_point(aes(color = pollster)) +
  geom_smooth(color = "blue") +
  theme(legend.position = "bottom")
```

```{r}
#| label: fig-pollster_trump
#| fig-cap: As election day draws close, pollsters are wrapping up their polls and posting their predictions on who will win. The polls in this graph predict Trump's win percentage, with the average in the blue line. The dots in the graph below represent individual polls conducted in the months leading up to the election. They are coloured based on the pollster who authored the poll.  
#| echo: false
#| warning: false
#| message: false

# Color by transparency_score
base_plot_trump +
  geom_point(aes(color = pollster)) +
  geom_smooth(color = "red") +
  theme(legend.position = "bottom")
```

### `states`

```{r}
#| label: fig-state
#| fig-cap: Battleground states are what usually determines the winner of a presidential election. This figure outlines the polling outcomes of each candidate from August to November, for each of the seven battleground states this election cycle.
#| echo: false
#| warning: false
#| message: false
#| 
# Define major swing states
swing_states <- c("Arizona", "Georgia", "Michigan", "Nevada", "North Carolina", "Pennsylvania", "Wisconsin")

combined_data <- bind_rows(
  just_harris_high_quality |> mutate(candidate = "Kamala Harris"),
  just_trump_high_quality |> mutate(candidate = "Donald Trump")
)

# Filter data for Harris and Trump in the swing states
swing_states_data <- combined_data |>
  filter(state %in% swing_states)

# Create a base plot for swing states data
swing_states_plot <- ggplot(swing_states_data, aes(x = end_date, y = pct, color = candidate)) +
  theme_minimal() +
  labs(x = "Polling End Date", y = "Polling Percentage", title = "Polling Outcomes in Swing States") +
  theme(legend.position = "bottom")

# Facet the plot by state
swing_states_plot +
  geom_point(aes(shape = candidate), alpha = 0.7) +
  geom_smooth(aes(linetype = candidate), se = FALSE) +
  facet_wrap(vars(state)) +
  scale_color_manual(values = c("Kamala Harris" = "blue", "Donald Trump" = "red"))

### `end_date`

Every poll has a start and end date for querying its participants. Polls that are conducted closer to election day may be more reflective of American opinions as opposed to polls conducted in August or September. This is why we will be using this variable as one of our predictors in the model.

```{r}
#| label: fig-enddate_harris
#| fig-cap: Polls are conducted from a start date to an end date. The closer the end date is to election day, the more insightful the results are to analysts and the general public. This distribution shows the end date of polls predicting a Kamala Harris win.
#| echo: false
#| warning: false
#| message: false

#### Plot data ####
just_harris_high_quality %>%
  ggplot(mapping=aes(x = end_date)) +
  geom_histogram(fill = "blue", alpha = 0.7, color = "black",)
```

```{r}
#| label: fig-enddate_trump
#| fig-cap: Polls are conducted from a start date to an end date. The closer the end date is to election day, the more insightful the results are to analysts and the general public. This distribution shows the end date of polls predicting a Donald Trump win.
#| echo: false
#| warning: false
#| message: false

#### Plot data ####
just_trump_high_quality %>%
  ggplot(mapping=aes(x = end_date)) +
  geom_histogram(fill = "red", alpha = 0.7, color = "black",)
```
